Solving Collective Spin Hamiltonians with Decoherence

### 2018-09-26 soc_squeezing/cumulant_expansions.tex
removed incorrect interpretation of a Stirling number
minor changes to text

We can expand operators in a Taylor series about $t=0$ to write
\begin{align}
  S_+^\ell S_\z^m S_-^n
  = \sum_{k\ge0} \f{t^k}{k!}
  \left.\f{d^k}{dt^k}\right|_{t=0} \p{S_+^\ell S_\z^m S_-^n},
\end{align}
which implies
\begin{align}
  \bk{S_+^\ell S_\z^m S_-^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{p,q,r} T^{\ell m n}_{pqr;k} \bk{S_+^p S_\z^q S_-^r}_{t=0},
  \label{eq:time_series}
\end{align}
where the matrix elements of the $k$-th time derivative are given by
\begin{align}
  T^{\ell m n}_{pqr;0}=\delta_{\ell p}\delta_{mq}\delta_{nr},
  &&
  T^{\ell m n}_{pqr;1}\equiv T^{\ell m n}_{pqr},
\end{align}
\begin{align}
  T^{\ell mn}_{pqr;k>1}
  \equiv \sum_{\v\lambda,\v\mu,\v\nu}
  T^{\ell mn}_{\lambda_1 \mu_1 \nu_1}
  T^{\lambda_1 \mu_1 \nu_1}_{\lambda_2 \mu_2 \nu_2}
  T^{\lambda_2 \mu_2 \nu_2}_{\lambda_3 \mu_3 \nu_3}
  \cdots T^{\lambda_{k-1} \mu_{k-1} \nu_{k-1}}_{pqr}.
\end{align}
Instead of explicitly choosing a weight function that determines which
operators to keep track of, we can now truncate the series in
\eqref{eq:time_series} by taking
\begin{align}
  \bk{S_+^\ell S_\z^m S_-^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{p,q,r} T^{\ell m n}_{pqr;k} \bk{S_+^p S_\z^q S_-^r}_{t=0}
  \to \sum_{k<M} \f{t^k}{k!}
  \sum_{p,q,r} T^{\ell m n}_{pqr;k} \bk{S_+^p S_\z^q S_-^r}_{t=0}.
\end{align}
Examining the nonzero entries of $T^{\ell mn}_{pqr;k}$ for
$k=0,1,\cdots,M-1$ now {\it tells} us which operators we need to keep
track of in order to compute the expectation value
$\bk{S_+^\ell S_\z^m S_-^n}$.  Furthermore, after computing the
expectation values of all relevant operators at $t=0$, time evolution
of $\bk{S_+^\ell S_\z^m S_-^n}$ takes only $\O\p{M}$ CPU time.
Finally, the relation
$\p{S_+^\ell S_\z^m S_-^n}^\dag=S_+^n S_\z^m S_-^\ell$ can be
exploited to cut both the number of expectation values
$\bk{S_+^p S_\z^r S_-^q}_{t=0}$ and the number of matrix elements
$T^{\ell mn}_{pqr}$ which we need to compute roughly in half.

@@

In the case of squeezing through spin-orbit coupling with the Sr-87
optical lattice clock, the primary form of decoherence is through
decay of individual spins, i.e.~with jump operators $\sigma_-^{(j)}$.
The two axis twisting Hamiltonians $H_{\t{TAT}}^{\z,\y}$ and
$H_{\t{TAT}}^{\z,\y}$, however, are actually realized in a rotating
frame, which implies that we must transform these jump operators
appropriately.  Specifically, the TAT protocol begins with a
time-dependent transverse-field Hamiltonian
\begin{align}
  H = \chi S_\z^2 + \Omega\p{t} S_\x,
  &&
  \Omega\p{t} = \beta\omega\cos\p{\omega t} S_\x,
\end{align}
where $\beta$ is the modulation index of the driving field, and the
drive frequency $\omega\gg N\chi$.  Moving into the rotating frame of
$\Omega\p{t}S_\x$ subtracts this term from the Hamiltonian, and
transforms operators $\O$ as
\begin{align}
  \O \to U\p{t}^\dag \O U\p{t},
  &&
  U\p{t} \equiv \exp\p{-i\int_0^t\d\tau~\Omega\p{\tau}S_\x}
  = \exp\sp{-i\beta\sin\p{\omega t}S_\x},
\end{align}
In particular, the operators $\tilde S_\pm\equiv-S_\z\pm iS_\y$
transform as
\begin{align}
  \tilde S_\pm
  \to U^\dag \tilde S_\pm U
  = e^{\pm i\beta\sin\p{\omega t}} \tilde S_\pm.
\end{align}
For large drive frequencies $\omega$, we can generally make a secular
approximation to say
\begin{align}
  e^{\pm i\beta\sin\p{\omega t}}
  = \sum_{n=-\infty}^\infty J_n\p{\pm\beta} e^{in\omega t}
  \approx J_0\p{\pm\beta} = J_0\p{\beta},
\end{align}
where $J_n$ is the $n$-th order Bessel function of the first kind.
Expanding $S_\z^2=\f14\p{\tilde S_++\tilde S_-}^2$, one can thus work
out that the effective Hamiltonian in the rotating frame of the drive
is
\begin{align}
  H_{\t{eff}}
  \approx \f{\chi}{2} \p{\sp{J_0\p{2\beta}+1} S_\z^2
  - \sp{J_0\p{2\beta}-1} S_\y^2},
\end{align}
where choosing $\beta$ such that $J_0\p{2\beta}=\pm1/3$ will give us
$H_{\t{TAT}}^{\z,\y}$ or $H_{\t{TAT}}^{\y,\x}$ up to irrelevant terms.
In a similar spirit, one can work out that
\begin{align}
  \tilde\sigma_\pm^{(j)}
  \equiv \f12\p{-\sigma_\z^{(j)}\pm i\sigma_\y^{(j)}}
  \to U^\dag \tilde\sigma_\pm^{(j)} U
  = e^{\pm i\beta\sin\p{\omega t}} \tilde\sigma_\pm^{(j)}
  \approx J_0\p{\beta} \tilde\sigma_\pm^{(j)},
\end{align}
which implies
\begin{align}
  \sigma_\z^{(j)} \to J_0\p{\beta} \sigma_\z^{(j)},
  &&
  \sigma_\pm^{(j)}
  \to \f12\sp{1 \pm J_0\p{\beta}} \sigma_+^{(j)}
  + \f12\sp{1 \mp J_0\p{\beta}} \sigma_-^{(j)}.
\end{align}
The TAT protocol about $\p{\z,\y}$ requires
\begin{align}
  J_0\p{2\beta} = 1/3
  \implies \beta \approx 0.906
  \implies J_0\p{\beta} \approx 0.805,
\end{align}
while the TAT protocol about $\p{\y,\x}$ requires
\begin{align}
  J_0\p{2\beta} = -1/3
  \implies
  \begin{cases}
    \beta \approx 1.63
    \implies J_0\p{\beta} \approx 0.440, \\
    \beta \approx 2.22
    \implies J_0\p{\beta} \approx 0.099.
  \end{cases}
\end{align}

@@

kind, i.e.~the number of ways to pick a set of $k$ elements from a set
of $n$ elements, and finally

>>>>

We can expand operators in a Taylor series about $t=0$ to write
\begin{align}
  S_+^\ell S_\z^m S_-^n
  = \sum_{k\ge0} \f{t^k}{k!}
  \left.\f{d^k}{dt^k}\right|_{t=0} \p{S_+^\ell S_\z^m S_-^n},
\end{align}
which implies
\begin{align}
  \bk{S_+^\ell S_\z^m S_-^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{p,q,r} T^{\ell m n}_{pqr;k} \bk{S_+^p S_\z^q S_-^r}_{t=0},
  \label{eq:time_series}
\end{align}
where the matrix elements of the $k$-th time derivative are given by
\begin{align}
  T^{\ell m n}_{pqr;0}=\delta_{\ell p}\delta_{mq}\delta_{nr},
  &&
  T^{\ell m n}_{pqr;1}\equiv T^{\ell m n}_{pqr},
\end{align}
\begin{align}
  T^{\ell mn}_{pqr;k>1}
  \equiv \sum_{\v\lambda,\v\mu,\v\nu}
  T^{\ell mn}_{\lambda_1 \mu_1 \nu_1}
  T^{\lambda_1 \mu_1 \nu_1}_{\lambda_2 \mu_2 \nu_2}
  T^{\lambda_2 \mu_2 \nu_2}_{\lambda_3 \mu_3 \nu_3}
  \cdots T^{\lambda_{k-1} \mu_{k-1} \nu_{k-1}}_{pqr}.
\end{align}
Instead of explicitly choosing a weight function that determines which
operators to keep track of, we can now truncate the series in
\eqref{eq:time_series} by taking
\begin{align}
  \bk{S_+^\ell S_\z^m S_-^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{p,q,r} T^{\ell m n}_{pqr;k} \bk{S_+^p S_\z^q S_-^r}_{t=0}
  \to \sum_{k<M} \f{t^k}{k!}
  \sum_{p,q,r} T^{\ell m n}_{pqr;k} \bk{S_+^p S_\z^q S_-^r}_{t=0}.
\end{align}
Examining the nonzero entries of $T^{\ell mn}_{pqr;k}$ for
$k=0,1,\cdots,M-1$ now {\it tells} us which operators we need to keep
track of in order to compute the expectation value
$\bk{S_+^\ell S_\z^m S_-^n}$.  Furthermore, after computing all
relevant matrix elements $T^{\ell mn}_{pqr;k}$ and expectation values
$\bk{S_+^p S_\z^q S_-^r}_{t=0}$, computing
$\bk{S_+^\ell S_\z^m S_-^n}$ takes only a short time independent of
system size.  Finally, the relation
$\p{S_+^\ell S_\z^m S_-^n}^\dag=S_+^n S_\z^m S_-^\ell$ can be
exploited to cut both the number of matrix elements
$T^{\ell mn}_{pqr}$ and expectation values
$\bk{S_+^p S_\z^r S_-^q}_{t=0}$ that we need to compute roughly in
half.

@@

In the case of squeezing through spin-orbit coupling with the Sr-87
optical lattice clock, the primary form of decoherence is through
decay of individual spins, i.e.~with jump operators $\sigma_-^{(j)}$.
The two axis twisting Hamiltonians $H_{\t{TAT}}^{\z,\y}$ and
$H_{\t{TAT}}^{\z,\y}$, however, are actually realized in a rotating
frame, which implies that we must transform these jump operators
appropriately.  Specifically, the TAT protocol begins with a
time-dependent transverse-field Hamiltonian
\begin{align}
  H = \chi S_\z^2 + \Omega\p{t} S_\x,
  &&
  \Omega\p{t} = \beta\omega \cos\p{\omega t},
\end{align}
where $\beta$ is the modulation index of the drive, and the drive
frequency $\omega\gg N\chi$, with $N$ the total number of spins.
Moving into the rotating frame of $\Omega\p{t}S_\x$ subtracts this
term from the Hamiltonian, and transforms operators $\O$ as
\begin{align}
  \O \to U\p{t}^\dag \O U\p{t},
  &&
  U\p{t} \equiv \exp\sp{-i\int_0^t\d\tau~\Omega\p{\tau}S_\x}
  = \exp\sp{-i\beta\sin\p{\omega t}S_\x},
\end{align}
In particular, the operators $\tilde S_\pm\equiv-S_\z\pm iS_\y$
transform simply as
\begin{align}
  \tilde S_\pm
  \to U^\dag \tilde S_\pm U
  = e^{\pm i\beta\sin\p{\omega t}} \tilde S_\pm.
\end{align}
For any operator $\O$ and drive frequency $\omega\gg\norm{\O}$, where
$\norm{\O}$ denotes the operator norm of $\O$, we can generally make a
secular approximation to say
\begin{align}
  e^{\pm im\beta\sin\p{\omega t}} \O
  = \sum_{n=-\infty}^\infty J_n\p{\pm m\beta} e^{in\omega t} \O
  \approx J_0\p{\pm m\beta} \O
  = J_0\p{m\beta} \O,
\end{align}
where $J_n$ is the $n$-th order Bessel function of the first kind.
Expanding $S_\z^2=\f14\p{\tilde S_+ + \tilde S_-}^2$, one can thus
work out that the effective Hamiltonian in the rotating frame of the
drive is
\begin{align}
  H_{\t{eff}}
  \approx \f{\chi}{2} \p{\sp{1+J_0\p{2\beta}} S_\z^2
    + \sp{1-J_0\p{2\beta}} S_\y^2}.
\end{align}
Driving with a modulation index $\beta$ for which
$J_0\p{2\beta}=\pm1/3$ then gives us the effective two-axis twisting
Hamiltonians
\begin{align}
  H_{\t{eff}}^{(+)}
  = \f{\chi}{3} \p{2 S_\z^2 + S_\y^2}
  \simeq \f{\chi}{3} \p{S_\z^2 - S_\x^2},
  &&
  H_{\t{eff}}^{(-)}
  = \f{\chi}{3} \p{S_\z^2 + 2 S_\y^2}
  \simeq \f{\chi}{3} \p{S_y^2 - S_\x^2},
\end{align}
where $\simeq$ denotes equality up to overall factors of
$\v S^2=S_\z^2+S_\x^2+S_\y^2$ which are irrelevant in the absence of
coherent coupling between states with different net spin.  In a
similar spirit, one can work out that single-spin operators transverse
to the $\x$-axis transform as
\begin{align}
  \tilde\sigma_\pm^{(j)}
  \equiv \f12\p{-\sigma_\z^{(j)}\pm i\sigma_\y^{(j)}}
  \to U^\dag \tilde\sigma_\pm^{(j)} U
  = e^{\pm i\beta\sin\p{\omega t}} \tilde\sigma_\pm^{(j)}
  \approx J_0\p{\beta} \tilde\sigma_\pm^{(j)},
\end{align}
which implies
\begin{align}
  \sigma_\z^{(j)} \to J_0\p{\beta} \sigma_\z^{(j)},
  &&
  \sigma_\pm^{(j)}
  \to \f12\sp{1 \pm J_0\p{\beta}} \sigma_+^{(j)}
  + \f12\sp{1 \mp J_0\p{\beta}} \sigma_-^{(j)}.
\end{align}
The TAT protocol about $\p{\z,\y}$ requires
\begin{align}
  J_0\p{2\beta} = 1/3
  \implies \beta \approx 0.906
  \implies J_0\p{\beta} \approx 0.805,
\end{align}
while the TAT protocol about $\p{\y,\x}$ requires
\begin{align}
  J_0\p{2\beta} = -1/3
  \implies
  \begin{cases}
    \beta \approx 1.63
    \implies J_0\p{\beta} \approx 0.440, \\
    \beta \approx 2.22
    \implies J_0\p{\beta} \approx 0.099.
  \end{cases}
\end{align}

@@

kind, and finally
